---
title: "Oomycete Sturgeon Eggs"
author: "Kristi Gdanetz MacCready"
date: "2022-05-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/")
```

# Load libraries
```{r}
library(vegan)
library(tidyverse)
library(phyloseq)
library(broom)
library(agricolae)
library(cowplot)
theme_set(theme_cowplot())
set.seed(333)
source("~/Documents/GitHub/DenefLab-MicrobeMiseq/R/miseqR.R")
```

# Set color palette
```{r}
my_colors3 <- c("#8ed5f0","#ffb2b2",
                "#bc93d1","#ffe564",
                "#d2fe4c","#ff7fbf",
                "#7f0000","#004c66",
                "#995100","#3c1451",
                "#660033", "#2c4c00",
                "#ff7f7f","#33b5e5",
                "#ffbb33","#aa66cc",
                "#ff3298", "#99cc00",
                "#cc0000","#007299",
                "#cc6c00","#6b238e",
                "#b20058", "#436500",
                "#ff4444","#0099cc",
                "#ff8800","#9933cc",
                "#669900","#e50072")

# Color codes for figures: Tube A = #ff7f7f, Tube B = #33b5e5, 
# Tube C = #ffbb33, Tube D = #aa66cc, Tube F = #99cc00
```

# Custom fuction
import indivual files, merge to create phyloseq object 
```{r}
import_mrg_physq = function(otu.table, 
                            map.file, 
                            tax.table, 
                            new.physq.obj){
  f.otu <- read.delim(otu.table, 
                     row.names=1, 
                     comment.char="!") %>% 
    as.matrix()  %>% 
    otu_table(taxa_are_rows = TRUE)
  
  f.mapping <- read.delim(map.file, 
                        row.names=1, 
                        comment.char="!") %>%
    as.data.frame() %>%
    sample_data()
  
  f.taxonomy <- read.delim(tax.table, 
                         header=TRUE, 
                         row.names=1) %>%
    as.matrix() %>%
    tax_table()
  
  new.physq.obj <- phyloseq(f.otu, f.mapping, f.taxonomy)
  return(new.physq.obj)
}
```

# Import and filter data
Sample metadata
```{r}
meta_oomy <- read_delim(file = "~/Documents/fish_eggs/oomy_sample_mapping_file.txt", delim = "\t")
```
Make phyloseq object
```{r}
physq.oomy <- import_mrg_physq("~/Documents/fish_eggs/oomy.otu_tab.txt",
                                "~/Documents/fish_eggs/oomy_sample_mapping_file.txt", 
                                "~/Documents/fish_eggs/edit_oomy_otus_taxonomy_05_sintax.txt", 
                                physq.oomy) #import 3 files need to create phyloseq object
```

# Summary statistics
Calculate summary statistics, check data distribution 
```{r}
ntaxa(physq.oomy) #495
nsamples(physq.oomy) #36

no.zero <- physq.oomy %>%
  prune_taxa(taxa_sums(.) > 5, .) %>%
  prune_samples(sample_sums(.) > 20, .) #416 taxa

sample_sum_df <- data.frame(sum = sample_sums(no.zero))
# Histogram of sample read counts
ggplot(sample_sum_df, aes(x = sum)) + 
  geom_histogram(color = "black", fill = "indianred", binwidth = 2500) +
  ggtitle("Distribution of sample sequencing depth") + 
  xlab("Read counts") +
  theme(axis.title.y = element_blank())

# summary statistics
smin <- min(sample_sums(no.zero)) #18,795
smean <- mean(sample_sums(no.zero)) #92,165
smax <- max(sample_sums(no.zero)) #145,784

# pull out taxonomy table for Krona figure
oomy.tax.clean <- tax_table(no.zero)
```

# Create barplot 
Format data 
```{r}
# normalization
nsample <- nsamples(no.zero)
sample_prop <- transform_sample_counts(no.zero, function(x) x/sum(x))
taxa_sums(sample_prop)/nsample # this gives you the abundances
sum(taxa_sums(sample_prop)/nsample) # this should be 1 for each sample
sample_sums(sample_prop) # this should be 1 too

# remove low abundance OTUs
no_low = filter_taxa(no.zero, function(x) mean(x) > 0.003, TRUE)
#no_low <- prune_taxa(taxa_sums(no.zero)>=100, no.zero)
ntaxa(no_low) #185 or 416
no_low = prune_taxa(taxa_sums(no_low) > 0, no_low)
nsamples(no_low) #36

# format for neat plotting
taxa <- no_low
glom <- tax_glom(taxa, "Genus") # merge taxa
C <- prune_taxa(names(sort(taxa_sums(glom), decreasing=T)), taxa)

# transform to abundance 
norm <- transform_sample_counts(C, function(x) x/sum(x)) 
colSums(otu_table(norm)[,1:5])
norm.merge <- tax_glom(norm, "Genus")

# transform to relative abundance 
norm.merge = transform_sample_counts(norm.merge, function(x) (100*x/sum(x))/9)
colSums(otu_table(norm.merge)[,1:5])
```

### Plot it 
```{r}
physq.oomy.f <- no.zero %>%
  tax_glom(taxrank = "Genus") %>%                     # agglomerate at phylum level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
  psmelt() %>%                                         # Melt to long format
  #filter(Abundance > 0.02) %>%                         # Filter out low abundance taxa
  arrange(Phylum)                                      # Sort data frame alphabetically by phylum

# Plot 
barplot_oomy <- physq.oomy.f %>%
  mutate(Tube = factor(Tube, ordered = TRUE,
         levels = c("A", "B", "C", "D"),
         labels = c("EBL", "LBL", "LBR", "FBR")) ) %>%
  ggplot(aes(x = Tube, y = Abundance, fill = Genus)) + 
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#ffb2b2", "#8ed5f0", "#ffe564", "#bc93d1" )) +
  scale_y_continuous(name = "Relative Abundance (% of total sequences)",
                      breaks = c(0, 2.25, 4.5, 6.75, 9), 
                     labels = c(0, 25, 50, 75, 100),
                     expand = c(0,0)) +
  xlab("Sample") + 
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 8),  
        legend.title = element_text(size = 10) )
barplot_oomy

#saveRDS(barplot_oomy, file = "~/Documents/fish_eggs/Figures/barplot_oomy.RDS")
```

# Alpha diversity 
Calculate indices, format data frames for statistics and plotting 
```{r}
# remove empty taxa
no.zero <- prune_taxa(taxa_sums(no.zero) > 0, no.zero)

# Calculate Diversity Indices
alpha.oomy <- estimate_richness(no.zero, split = TRUE, 
                                measures = c("Observed", "Shannon"))

# add sample metadata to diversity indices
alpha.oomy2 <- alpha.oomy %>%
  rownames_to_column(var = "Sample_ID") %>%
  left_join(meta_oomy, by = "Sample_ID") %>%
  # format for plotting 
  pivot_longer(cols = c("Observed", "Shannon"),
               values_to = "Measure",
               names_to = "Index") %>% 
  # relabel samples 
  mutate(Tube = factor(Tube, ordered = TRUE,
         levels = c("A", "B", "C", "D"),
         labels = c("EBL", "LBL", "LBR", "FBR")) ) %>%
  arrange(Tube, Index)

# summary statistics for plots
alpha.oomy.sum <- alpha.oomy2 %>%
  arrange(Tube, Index) %>% group_by(Tube, Index) %>%
  summarise(Mean_index = mean(Measure, na.rm = TRUE),
            StDev_index = sd(Measure, na.rm = TRUE)) 
```

### Tests 
test if diversity indices are different
```{r}
### Shannon index
ANOVA1 <- alpha.oomy2 %>%
  filter(Index == "Shannon") %>%
  aov(Measure~Tube, data = .)
summary(ANOVA1) #p = 2.62e-05

# conduct Tukey"s HSD test with agricolae and get letters for figure 
letters1 <- as.matrix(HSD.test(ANOVA1,"Tube", group=TRUE, alpha = 0.05)$groups) %>%
  as.data.frame() %>% #formatting the table with the letters
  rownames_to_column(var="Tube") %>%
  mutate(Measure = round(as.numeric(as.character(Measure)), digits=2),
         groups = as.character(groups)) %>%
  rename(Shannon = Measure)

### observed ASVs index
ANOVA2 <- alpha.oomy2 %>%
  filter(Index == "Observed") %>%
  aov(Measure~Tube, data = .)
summary(ANOVA2) #p = 0.671
letters2 <- as.matrix(HSD.test(ANOVA2,"Tube", group=TRUE, alpha = 0.05)$groups) %>%
  as.data.frame() %>% #formatting the table with the letters
  rownames_to_column(var="Tube") %>%
  mutate(Measure = round(as.numeric(as.character(Measure)), digits=0),
         groups = as.character(groups)) %>%
  rename(Observed = Measure)
```

### Plot
```{r}
# generate plot
richplot_oomy <- alpha.oomy2 %>%
  ggplot(aes(x = Tube, y = Measure, color = Tube, Index)) +
  facet_wrap(.~Index, scales = "free") +
  # plot means and error bars
  geom_errorbar(data = alpha.oomy.sum, inherit.aes = FALSE, width = 0.5, color = "black",
                aes(x = Tube, ymin = Mean_index - StDev_index,
                    ymax = Mean_index + StDev_index)) +
  geom_crossbar(data = alpha.oomy.sum, inherit.aes = FALSE, fatten = 2, 
                aes(x = Tube, color = Tube,
                    y = Mean_index, ymin = Mean_index, ymax = Mean_index )) +
  # plot individual data points 
  geom_jitter(width = 0.25) +
  # set colors 
  scale_color_manual(name = "Sample", values = c("#ff7f7f", "#33b5e5", "#ffbb33", "#aa66cc", "#99cc00")) +
  # format axes
  ylab("Alpha Diversity Measure") + xlab("Sample") +
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 10))
richplot_oomy
```

# Beta diversity 
```{r}
# scale to even depth
dat_scale <- no.zero %>%
  scale_reads(round = "round") 
# repeat summary statistics
smin <- min(sample_sums(dat_scale)) #18,789
smean <- mean(sample_sums(dat_scale)) #18,794
smax <- max(sample_sums(dat_scale)) #18,799

# ordinate
dat_nmds <- ordinate(
  physeq = dat_scale, 
  method = "NMDS", k=3, maxit=500, try=100,
  distance = "bray")
dat_nmds #stress=0.08947188 
```

### Plot
```{r}
# create dotplot 
nmds_oomy <- plot_ordination(
  physeq = dat_scale,
  ordination = dat_nmds,
  color = "Tube") + 
  scale_color_manual(name = "Sample", values = c("#ff7f7f", "#33b5e5", "#ffbb33", "#aa66cc")) +
  geom_point(aes(color = Tube), size = 4) +
  # format axes
  theme(axis.text = element_text(size = 8),
        axis.title = element_text(size = 10),
        legend.text = element_text(size = 8),  
        legend.title = element_text(size = 10) )
nmds_oomy
```

```{r}
# Permanova and Betadisp 
dat_bray <- phyloseq::distance(dat_scale, method="bray") #calculate bray curtis distance matrix
# make a data frame from the sample_data
sampledf <- data.frame(sample_data(no.zero))

# Adonis test 
adon1 <- adonis(dat_bray~Tube, data=sampledf)
adon1 #p=0.001

# Homogeneity of dispersion test 
beta <- betadisper(dat_bray, sampledf$Tube)
permutest(beta) #p=0.004

# Anosim test 
ano_group <- get_variable(dat_scale, "Tube")
ano <- anosim(dat_bray, ano_group)
ano$signif #p=0.001
ano$statistic #R=0.5766604
```

# Export Figures
Diversity plots 
```{r}
figure2 <- plot_grid(richplot_oomy + theme(legend.position = "none"), 
                     nmds_oomy + theme(legend.position = "none"),
                     nrow = 1, ncol = 2, 
                     labels = c("A", "B"))
figure2
ggsave2(filename = "~/Documents/fish_eggs/Figures/figure2.svg", figure2,
        device = "svg", dpi = 300, width = 178, height = 100, units = "mm")
```

Supplemental table
```{r}
# Tax table
oomy.tax.clean <- tax_table(no.zero) %>%
  as.data.frame() %>%
  rownames_to_column(var = "OTU")
# Shared table
oomy.otu.clean <- otu_table(no.zero) %>%
  as.data.frame() %>%
  rownames_to_column(var = "OTU")
# Combine
supp_tab1 <- inner_join(oomy.tax.clean, oomy.otu.clean, by = "OTU") %>%
  rename("EBL_1-1" = "KSOOA11", "EBL_1-2" = "KSOOA12", "EBL_1-3" = "KSOOA13",  
         "EBL_2-1" ="KSOOA21", "EBL_2-2" = "KSOOA22", "EBL_2-3" ="KSOOA23", 
         "EBL_3-1" ="KSOOA31", "EBL_3-2" = "KSOOA32", "EBL_3-3" ="KSOOA33", 
         "LBL_1-1" = "KSOOB11", "LBL_1-2" = "KSOOB12", "LBL_1-3" = "KSOOB13", 
         "LBL_2-1" = "KSOOB21", "LBL_2-2" = "KSOOB22", "LBL_2-3" = "KSOOB23",
         "LBL_3-1" = "KSOOB31", "LBL_3-2" = "KSOOB32", "LBL_3-3" = "KSOOB33",
         "LBR_1-1" = "KSOOC11", "LBR_1-2" = "KSOOC12", "LBR_1-3" = "KSOOC13",
         "LBR_2-1" = "KSOOC21", "LBR_2-2" = "KSOOC22", "LBR_2-3" = "KSOOC23",
          "LBR_3-1" = "KSOOC31", "LBR_3-2" = "KSOOC32", "LBR_3-3" = "KSOOC33",
         "FBR_1-1" = "KSOOD11", "FBR_1-2" = "KSOOD12", "FBR_1-3" = "KSOOD13", 
         "FBR_2-1" = "KSOOD21", "FBR_2-2" = "KSOOD22", "FBR_2-3" = "KSOOD23", 
         "FBR_3-1" = "KSOOD31", "FBR_3-2" = "KSOOD32", "FBR_3-3" = "KSOOD33"  )
write_delim(supp_tab1, file = "~/Documents/Supplemental_Table1.txt",
            delim = "\t", col_names = TRUE, na = "NA")
```

-----
end 
